Deep Multimodal Generative Models for Weakly-Supervised Learning

### 2017-11-17 paper/draft.tex
Rewrite the intro+related works

\section{Introduction}
The problem of learning from multiple modalities is especially important since the majority of modern information is represented through multiple channels. For example, images in the web are often embedded around text. And while images can be described by pixels, we expect related modalities like text to also hold relevant features. People often interact with this multimodal information in a \textit{bi-directional} way. For example, one can imagine what a black cat looks like but also prescribe this description to a photo. Additionally, available multimodal information is often \textit{sparse}, in the sense there exist many large corpuses of images or text alone but few datasets exist (and are expensive to construct) that contain multiple modalities together. Ideally, we would like the generative models we build to extract a joint representation that captures high-level concepts across modalities in a way that is bi-directional and weakly-supervised.

In this work, we propose a multimodal variational autoencoder (MMVAE). In our graphical model, we have two modalities $x_{1}$ and $x_{2}$ that are each conditioned on a latent variable $z$, which models the joint distribution, $p(x_{1}, x_{2})$. Such a modal is bi-directional since we can reconstruct both modalities from either $x_{1}$ or $x_{2}$. We will also show that we can train the MMVAE with sparse $(x_{1}, x_{2})$ pairs, and heavily utilize on a larger set of independent $(x_{1})$ and $(x_{2})$ examples. The most significant feature of MMVAEs is that it does not create a separate multimodal encoder $q(z | x_{1}, x_{2})$; instead it shares parameters with the encoders from each modality, $q(z | x_{1})$, $q(z | x_{2})$. This has the benefit of producing a simpler model while ensuring that high-dimensional missing modalities do not cause our generated samples to collapse.

\subsection{Related Work}
In deep learning, multimodal learning is often done using separate branches in the neural net for each modality, and a common top hidden layer. For instance, \citet{ngiam2011multimodal} trained deep autoencoders on audio and video input and found that the bimodal representations were better than any single equivalent. More recently, VAEs \cite{kingma2013auto, kingma2014semi} have been used to train models in high-dimensional multimodal settings. For example, conditional VAEs \cite{sohn2015learning} maximize a conditional log-likleihood by variational methods where often one modality is conditioned on another i.e. handwriting digits and labels, faces and attributes, or captioning. Notably, conditional VAEs (CVAE) are not bi-directional.

\citet{pandey2017variational} proposed the conditional multimodal autoencoder (CMMA), which also maximizes a conditional log-likelihood. CMMA explicitly connects the conditional variable the latent variable, where CVAE does not. Moreover, it also constrains the latent representation to be close to the joint representation across modalities. However, CMMA is still not bi-directional.

Finally, \citet{suzuki2016joint} proposed a joint multimodal VAE (JMVAE). This is similar to our work in that it is bi-directional, handles missing modalities, and models the joint distribution. Our proposed model (MMVAE) and JMVAE share the same graphical notation, but differ in our approach. Crucially, JMVAE trains separate encoders for each modality and an additional one for the multimodal case. It then includes a KL divergence term \textit{per modality} in the loss. In constrast, MMVAE does not have an additional encoder for the multimodal case and does not have additional KL terms. It does so by sharing encoder parameters to learn the guide distributions: $q_{\phi_{x_{1}}}(z|x_{1})$, $q_{\phi_{x_{2}}}(z|x_{2})$, $q_{\phi}(z|x_{1}, x_{2})$

>>>>

\section{Introduction}
The problem of learning from multiple modalities is especially important since the majority of modern information is represented through multiple channels. For example, images in the web are often embedded around text. And while images can be described by pixels, we expect related modalities like text to also hold relevant features. People often interact with this multimodal information in a \textit{bi-directional} way. For example, one can imagine what a black cat looks like but also prescribe this description to a photo. Additionally, available multimodal information is often \textit{sparse}. There exist many large corpuses of images or text alone but few datasets exist (and are expensive to construct) that contain multiple modalities together. Ideally, we would like the generative models we build to extract a joint representation that captures high-level concepts across modalities in a way that is bi-directional and weakly-supervised.

In this work, we propose a multimodal variational autoencoder (MMVAE), an extension of the traditional VAE framework to handle multiple input types. The first extension is model the setting where we have both an image, $x$, and a natural text, $y$. We then assume a joint generative model of the form $p(x, y, z) = p(z)p(x | z)p(y | z)$ where $p(z)$ is the prior over the latent variable $z$, $p(x | z)$ is our image decoder, and $p(y | z)$ is our text decoder. We then further extend the VAE using a novel objective function, which we call the \textit{MMELBO}, for simultaneously training the model from paired data, $D_{p} = \{ (x_{n}, y_{n}) \}$, and unimodal data, $D_{x} = \{ x_{n} \}$, $D_{y} = \{ y_{n} \}$. Namely, using MMELBO allows for both evaluating unpaired data at test time and training with sparse paired examples.

MMVAE jointly fits three encoders: $q(z | x, y)$, $q(z | x)$, and $q(z | y)$. Doing so, we can project images and natural text into the same latent space, and offer ways to translate between the two (bi-directional). For efficiency and simplicity, MMVAEs share parameters between the three encoders. The encoders interact via product of experts (POE) \cite{hinton2006training, vedantam2017generative}.

We report experimental results on three datasets. The first dataset is MNIST, where labels are stringified to text as the second modality. The second dataset is MultiMNIST, where 0 to 4 MNIST digits are placed on a 50x50 canvas. Unlike the traditional MultiMNIST, we do not vary the locations of these 0 to 4 digits. We again use the stringified label as a the second modality, where the label is the digit class from top-left to bottom-right. We briefly discuss introducing randomness into the text by permuting the string. The third dataset is the Microsoft COCO Captions dataset \cite{chen2015microsoft}, which contains 330K images and 5 captions per image. We show that our model performs competitively on these datasets.

Finally, for each of the three datasets, we investigate how our model performs under weak-supervision by aggressively reducing the number of paired data points we provide to the model. We show that MMVAEs do not need a significant number of paired data points to learn a good joint representation across modalities, making this model feasible for real-world tasks i.e. caption generation or speech-to-text. We also briefly discuss reducing the number of data points of a single modality and its effect of learning.

In summary, the contributions of this paper are as follows: first, we present MMVAEs, introducing the MMELBO objective. Second, we show that the product of experts techniques works with two unstructured modalities: images and raw text. Lastly, we should that multimodal VAEs are good candidates for weakly supervised learning.

\subsection{Related Work}
In deep learning, multimodal learning is often done using separate branches in the neural net for each modality, and a common top hidden layer. For instance, \citet{ngiam2011multimodal} trained deep autoencoders on audio and video input and found that the bimodal representations were better than any single equivalent. More recently, VAEs \cite{kingma2013auto, kingma2014semi} have been used to train conditional generative models of the form $p(y | x)$ in high-dimensional bimodal settings. Here, $y$ is often a class label, a sentence, or another image depending on the task. For example, conditional VAEs \cite{sohn2015learning}, and conditional multimodal autoencoders (CMMA) \cite{pandey2017variational} maximize a conditional log-likelihood where often one modality is conditioned on another i.e. handwriting digits and labels, faces and attributes, etc. Notably, these CVAEs and CMMAs are not bi-directional. We are more interested in learning a shared latent space of $p(x, y)$ where we can condition interchangeably.

Recently, there have been several works on such joint models, many of which share commonalities to our proposed model. For example, the BiVCCA employs a similar VAE model with the objective shown below. It includes two inference networks, one for each modality. We note that MMVAE uses a third multimodal objective; we show in our experiments that doing so is important in learning a useful latent space. The BiVCCA loss encourages proper reconstruction of both $x$ and $y$ given only one; $\mu$ acts as a weight to prioritize different modalities.

\begin{equation}
    J(x, y, \theta, \phi) = \mu (E_{q_{\phi_{x}}(z|x)}[\textup{log}(p_{\theta_{x}}(x|z)) + \lambda \textup{ log}(p_{\theta_{y}}(y|z))] - \textup{KL}(q_{\theta_{x}}(z|x)), p_{\theta}(z)) + (1 - \mu) (E_{q_{\phi_{y}}(z|x)}[\textup{log}(p_{\theta_{x}}(x|z)) + \lambda \textup{ log}(p_{\theta_{y}}(y|z))] - \textup{KL}(q_{\theta_{y}}(z|y)), p_{\theta}(z))
\end{equation}

\citet{suzuki2016joint} then build onto the BiVCCA objective with their JMVAE model, that adds an additional multimodal encoder $q(z|x,y)$ and two additional KL divergence terms to bring $q(z|x,y)$ close to $q(z|x)$ and $q(z|y)$. More explicitly, their objective is the following:

\begin{equation}
    J(x, y, \theta, \phi) = \textup{elbo}_{1, \lambda, 1}(x, y, \theta, \phi) - \alpha[\textup{KL}(q_{\phi}(z|x,y), q_{\phi_{y}}(z|y)) + \textup{KL}(q_{\phi}(z|x,y), q_{\phi_{x}}(z|x))]
\end{equation}

As \citet{vedantam2017generative} note, forcing $q_{\phi(z|y)}$ to be close to $q_{\phi(z|x, y)}$ is somewhat unintuitive but if you rewrite $q_{\phi(z|x, y)}$ as $q_{\phi}^{\textup{avg}}(z|y)$, they show that it has the benefit of $q_{\phi(z|y)}$ covering the embeddings of all $x$ associated with $y$. \citet{vedantam2017generative} go on to introduce their own multimodal VAE objective, the \textit{TELBO} to achieve a similar effect with a simpler design. The TELBO is defined as

\begin{equation}
    J(x, y, \theta_{x}, \theta_{y}, \phi, \phi_{x}, \phi_{y}) = \textup{elbo}_{1, \lambda, 1}(x, y, \theta_{x}, \theta_{y}, \phi) + \textup{elbo}_{1, 1}(x, \theta_{x}, \phi_{x}) + \textup{elbo}_{\gamma, 1}(y, \theta_{y}, \phi_{y})
\end{equation}

where $\lambda, \gamma$ are hyperparameters of the ELBO. See model section for notation. The authors report training this model in two stages: first, using only aligned data, train on first ELBO term alone. Then, freezing the decoders $p_{\theta_{x}}(x|z)$ and $p_{\theta_{y}}(y|z)$, train on the last two ELBO terms above. The TELBO is very similar to our proposed objective (MMELBO). We denote two main differences. First, in TELBO, when the input is a single modality, the objective decomposes to the traditional ELBO, where it penalizes differences between the reconstruction and original of that modality only. In MMELBO, all three ELBO terms include reconstruction costs for all modalities. For example, given an unpaired image $x$, MELBO encourages its latent representation to decode to both a sensible image and text. We later show that this change is important in produce sharper samples. Second, training MMELBO is done concurrently in a single stage. We show paired and unpaired examples together and share encoder and decoder parameters across the three ELBO terms.
